Machine Learning guided workflow for Ribosome Binding Site engineering

### 2021-06-02 paper/main.tex
address colin's comments

However, de-novo design of small genetic elements is still challenging due to unknown relationships between sequence and performance of such elements. 
This means that many designers have to rely on known and characterised parts that may not be optimal for their constructs.
The problem with this approach is that such part libraries can also be unreliable due to poor reliability of methods used to obtain them.\\
The biggest limitation for the DBTL approach currently is the Learn part of the cycle - there is very limited access to methods and software that can improve and understand designs based on the experimental results.

@@

New experimental methods heavily relying on advances in automation and microfludics allow unprecedented precision and throughput in data generation.
These new datasets can be combined with data reliant machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018,Radivojevic2020}. 
In the past few years there was a significant uptake of machine learning based approaches in synthetic biology \cite{LAWSON2021}.
Jervis \emph{et al.} used support vector machines and neural network to optimise production of monoterpenoid in \emph{Esherichia coli} \cite{Jervis2019}.

@@

There were also successful attempts at using deep learning techniques for analysis of big datasets \cite{Alipanahi2015,Angermueller2016}.\\
However, the use of machine learning in synthetic biology is still in its infancy and will require additional research to show its full potential. 
The challenges of applying machine learning techniques on synthetic biology have two folds.
On the one hand, there is no large-scale and high-quality labelled data. 
The available RBS library \cite{jervis2018machine} only covers 56 unique labelled RBS.
And it is time-consuming and expensive to measure the TIR for RBS sequences. 
On the other hand,  while the goal is to find highest TIR among the design space, one also need to explore as much as possible to learn the whole design space to avoid getting stuck in the local maximum.  
We addressed the challenges by applying DBTL workflow. 
Instead of querying labels for all RBS sequences in the design space (4,138 in total), we recommend RBS sequences to query in batches and learn the prediction model in an online form.
Particularly, we use bandits algorithm for recommending RBS sequences to address the exploitation-exploration balance. \\
We present how machine learning algorithms can be used as part of the DBTL cycle to predict (Learn) and recommend (Design) variants of RBS with the goal of optimising associated protein expression level. 

@@

Our overall experimental goal was to maximise the Translation Initiation Rate (TIR) by querying the batches of RBS sequences while minimising the number of DBTL cycle turns that we had to do.

@@

In the zeroth round, randomised RBS sequences and preliminary machine learning recommended designs based on literature were designed to explore the experimental space. 
In the subsequent rounds, designs were recommended by the algorithm based on the data obtained in the previous rounds. 
The designs were then physically constructed in batches of 90 to fit our automated process (see \textbf{Methods} section).
After constructing, the plasmids harbouring the new genetic devices were tested in microplate reader.
The results were then fed back to the algorithm for it to recommend the next round of designs.

@@

In this work, we have shown how machine learning and high-throughput, automated laboratory methods can be used to efficiently generate a library of small parts, in this case bacterial RBS. 
We have used Gaussian Process regression to predict the shape of our function and Upper Confidence Bound Bandit algorithm to recommend sequences to be tested.
We have investigated a number of methods of digitising the DNA sequence, finally settling on Weighted Degree Kernel with Shift method, which fit well into our prediction method.
We building and testing, we have performed bulk of our experiments using automation to increase their speed, reliability and reproducibility.
By using our workflow, we have found very strong RBS sequences and we have generated an extensive library of diverse RBS that can be used in the future studies.
We have achieved this despite the relatively low accuracy of our predictions, which means that the presented algorithms are robust and able to identify the right signal in a noisy environment. \\

There are still open questions that need to be addressed for applying machine learning in synthetic biology.
For example, how can we understand the language of RBS sequences better?
In other words, can we extract more biologically important information from the decisions made by our algorithms?
Another example would be - given a small amount of RBS sequences tested, how can machine learning algorithm provide more accurate predictions and uncertainty measurement? 
Finally, another question with high impact on the performance of the workflow is - how should we choose the parameters which controls the exploration-exploitation rate? \\

@@

All code and data required to reproduce the results is available at Github: \url{https://github.com/mholowko/SynbioML}

>>>>

However, de-novo design of small genetic elements is still challenging due to unknown relationships between sequence and performance of such elements.\\
The biggest gap in the DTBL cycle, at present, is at the Learn and Design interface - it is hard to translate obtained results into new designs.

@@

These new data sets can be combined with data reliant machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018,Radivojevic2020,LAWSON2021}. 
For example, Jervis \emph{et al.} used support vector machines and neural network to optimise production of monoterpenoid in \emph{Esherichia coli} \cite{Jervis2019}.

@@

There were also successful attempts at using deep learning techniques for analysis of big synthetic biology data sets \cite{Alipanahi2015,Angermueller2016}.\\
In this work, we present how machine learning algorithms can be used as part of the DBTL cycle to predict (Learn) and recommend (Design) variants of bacterial RBS with the goal of optimising associated protein expression level. 

@@

Our overall experimental goal was to maximise the Translation Initiation Rate (TIR) by building and testing batches of RBS sequences while minimising the number of DBTL cycle turns that we had to do.

@@

@@

In this work, we have shown how machine learning and high-throughput, automated laboratory methods can be used to efficiently generate a library of small parts, in this case bacterial RBS. 
We have used Gaussian Process regression to predict the shape of our function and Upper Confidence Bound Bandit algorithm to recommend sequences to be tested.
We have investigated a number of methods of digitising the DNA sequence, finally settling on Weighted Degree Kernel with Shift method, which fit well into our prediction method.
In building and testing, we have performed bulk of our experiments using automation to increase their speed, reliability and reproducibility.
By using our workflow, we have found very strong RBS sequences and we have generated an extensive library of diverse RBS that can be used in the future studies.
We have achieved this despite the relatively low accuracy of our predictions, which means that the presented algorithms are robust and able to identify the right signal in a noisy environment. \\

There are still open questions that need to be addressed for applying machine learning in synthetic biology.
For example, how can we understand the language of RBS sequences better?
In other words, can we extract more biologically important information from the decisions made by our algorithms?
Another example would be - given a small amount of RBS sequences tested, how can machine learning algorithm provide more accurate predictions and uncertainty measurement? 
Next question would be - how generalizable is this method? Can it be succesfully used for design of bigger genetic parts, like promoters?
Finally, another question with high impact on the performance of the workflow is - how should we choose the parameters which controls the exploration-exploitation rate? \\

@@

All code and data required to reproduce the results is available at Github: \url{https://github.com/mholowko/SynbioML} .
All the processed and raw data is included in the repository.
Sequences of plasmids and oligos

### 2021-06-02 paper/methods.tex
address colin's comments

\textbf{\textit{E. coli} transformation.}
The DH5$\alpha$ cell line (Thermo Fisher Scientific, catalogue no. 18265017) was made chemically competent using the Mix $\&$ Go \textit{E. coli} Transformation Kit $\&$ Buffer Set (Zymo Research, catalogue no. T3001).
20$\mu$L of cells was aliquoted into each well of a cold 96-well PCR plate and stored at -80$^{\circ}$C for later use.
Plates of cells were thawed on a -20$^{\circ}$C cold block before 3$\mu$L of the assembly product was added and mixed using the CyBio FeliX liquid handler.
Cells were incubated on a cold block for 2-5 minutes before being plated in a 96 square grid on Omnitrays containing LB (BD, catalogue no. \mengyan{***}) with 34$\mu$g/mL chloramphenicol (Sigma, catalogue no. \mengyan{***}).
Plates were incubated overnight at 37$^{\circ}$C. 
8

@@

\textbf{Glycerol stock preparation.}
100$\mu$L of sterile 80\% (v/v) glycerol and 100$\mu$L of overnight culture were combined in the wells of a 96 deep (2mL) round well plate using the CyBio Felix liquid handler.
They were then sealed with a 96-well silicon sealing mat and transferred to a -80$^{\circ}$C freezer.

\textbf{Sequencing}
Strains that gave GFP fluorescence intensity readings similar to that of the original RBS were selected for sequence confirmation by capillary electrophoresis sequencing (CES) by Macrogen, Inc. (Seoul, South Korea).
The strains transformed with each of the selected constructs were grown to saturation in 5 mL LB medium with chloramphenicol selection (34 $\mu$g/mL).
Plasmids were extracted from the cultures using the QIAprep Spin Miniprep Kit (QIAGEN) according to the manufacturer's instructions.
Plasmid concentrations were quantified using the Cytation 5 plate reader with the Take3 Micro-Volume Plate (BioTek) and all fell in the range of 100-200 ng/$\mu$L.
Samples of 20 $\mu$L of undiluted plasmid DNA were sequenced using a single primer (5'-CGATATAGGCGCCAGCAA-3') that binds approximately 150 bp upstream of the RBS.
Reads were aligned with the template sequence in the Teselagen software (TeselaGen Biotechnology, Inc.).

@@

\textbf{Test strain culture.}
Overnight cultures were started by inoculating 1mL of LB medium supplemented with 34$\mu$g/mL chloramphenicol with ~2 $\mu$L of the glycerol stock in a 96 deep (2mL) round well plate.
Cultures were incubated at 37$^{\circ}$C with shaking (~300rpm) for ~17 hours. 
The following morning, 20$\mu$L of overnight cultures were added to 980$\mu$L of fresh selection medium and these cultures were grown at 37$^{\circ}$C with shaking in 2mL round well 96 plate. 
After 90 minutes, 200$\mu$L of each culture (induced with 1.0$\mu$L of 0.1M IPTG) was transferred to a flat-bottom clear polystyrene 96-well plate.

>>>>

\textbf{\textit{E. coli} transformation.}
The DH5$\alpha$ cell line (Thermo Fisher Scientific, catalogue no. 18265017) was made chemically competent using the Mix $\&$ Go \textit{E. coli} Transformation Kit $\&$ Buffer Set (Zymo Research, catalogue no. T3001).
20$\mu$L of cells was aliquoted into each well of a cold 96-well PCR plate and stored at -80$^{\circ}$C for later use.
Plates of cells were thawed on a -20$^{\circ}$C cold block before 3$\mu$L of the assembly product was added and mixed using the CyBio FeliX liquid handler.
Cells were incubated on a cold block for 2-5 minutes before being plated in a 96 square grid on Omnitrays containing LB (BD, catalogue no. 244610) and 15g/L agar (Sigma-Adrich, catalogue no. A1296) with 34$\mu$g/mL chloramphenicol (Sigma-Aldrich, catalogue no. C1919).
Plates were incubated overnight at 37$^{\circ}$C. 
8

@@

\textbf{Glycerol stock preparation.}
100$\mu$L of sterile 80\% (v/v) glycerol (Chem-Supply, catalogue no. GA010) and 100$\mu$L of overnight culture were combined in the wells of a 96 deep (2mL) round well plate using the CyBio Felix liquid handler.
They were then sealed with a 96-well silicon sealing mat and transferred to a -80$^{\circ}$C freezer.

\textbf{Sequencing}
Strains that gave GFP fluorescence intensity readings similar to that of the original RBS were selected for sequence confirmation by capillary electrophoresis sequencing (CES) by Macrogen, Inc. (Seoul, South Korea).
The strains transformed with each of the selected constructs were grown to saturation in 5 mL LB medium with chloramphenicol selection (34 $\mu$g/mL).
Plasmids were extracted from the cultures using the QIAprep Spin Miniprep Kit (QIAGEN, catalogue no. 27106) according to the manufacturer's instructions.
Plasmid concentrations were quantified using the Cytation 5 plate reader with the Take3 Micro-Volume Plate (BioTek) and all fell in the range of 100-200 ng/$\mu$L.
Samples of 20 $\mu$L of undiluted plasmid DNA were sequenced using a single primer (5'-CGATATAGGCGCCAGCAA-3') that binds approximately 150 bp upstream of the RBS.
Reads were aligned with the template sequence in the Teselagen software (TeselaGen Biotechnology, Inc.).

@@

\textbf{Test strain culture.}
Overnight cultures were started by inoculating 1mL of LB medium supplemented with 34$\mu$g/mL chloramphenicol with ~2 $\mu$L of the glycerol stock in a 96 deep (2mL) round well plate.
Cultures were incubated at 37$^{\circ}$C with shaking (~300rpm) for ~17 hours. 
The following morning, 20$\mu$L of overnight cultures were added to 980$\mu$L of fresh selection medium and these cultures were grown at 37$^{\circ}$C with shaking in 2mL round well 96 plate. 
After 90 minutes, 300$\mu$L of each culture (induced with 1.0$\mu$L of 0.1M IPTG (Sigma-Aldrich, catalogue no. I5502) to a final concentration of 0.5 mM) was transferred to a flat-bottom clear polystyrene 96-well plate.
